---
title: "Correlation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicrobiomeDB)
```

## What is a Correlation Analysis?

Correlations are useful for identifying relationships between variables. They are very helpful
for identifying biomarkers and functional associations, among other things.

Biomarkers in microbiome data are microbial taxa or features that exhibit a significant correlation
with particular conditions, traits, or variables in sample metadata. Identifying these biomarkers
allows researchers to uncover associations between the microbiome and external factors, such as age, 
disease status, or environmental conditions. Biomarkers serve as indicators of specific biological 
or clinical phenomena within microbial communities.

Functional associations in microbiome data refer to relationships between taxonomic relative abundances 
and abundances of specific pathways or other functional features. Understanding these associations 
allows researchers to explore how the taxonomic composition of microbial communities influences 
functional capabilities.

## Why Care About Biomarker Discovery?

Researchers are interested in biomarker discovery for several reasons:

**Diagnostic Insights**: Biomarkers can serve as potential diagnostic indicators, helping identify 
microbial patterns associated with specific conditions.

**Predictive Modeling**: Understanding biomarker correlations enables predictive modeling of microbial 
responses to external factors.

**Biological Significance**: Biomarkers provide insights into the biological significance of microbial 
community variations in response to different conditions.

## Why Care About Functional Associations?

**Biological Insights**: Explore how changes in taxonomic composition may impact the functional potential 
of microbial communities.

**Pathway-Level Analysis**: Understand how specific pathways or functional features correlate with 
taxonomic abundance, providing pathway-level insights.

**Predictive Modeling**: Assessing functional associations aids in predicting microbial functional 
responses to environmental changes or perturbations.

## How are Correlations Calculated?

This package employs correlation analysis between microbial taxonomic abundances and sample metadata 
or abundances of pathways or other functional data using the following approach:

Users can choose either Spearman or Pearson correlation for the analysis. Both will produce
a correlation coefficient and a p-value indicating statistical significance.

### Spearman Correlation

Use when the relationship between variables is monotonic but not necessarily linear. Suitable for 
non-linear associations.

```{r}
## first lets find some interesting data
microbiomeData::getCuratedDatasetNames()
getCollectionNames(microbiomeData::HMP_WGS)

## grab a collection of interest
HMP_WGS_species <- getCollection(microbiomeData::HMP_WGS, "WGS Species")

## get a correlation ComputeResult
## this is not necessarily to recommend spearman for metadata. 
## it is simply exemplary. Always look at your data!
species_vs_metadata <- correlation(HMP_WGS_species, method = 'spearman') 
```

### Pearson Correlation

Use when the relationship between variables is linear. Suitable for assessing linear associations.

```{r}
## grab two collections of interest, in this case species level data and pathway abundance data
HMP_WGS_species <- getCollection(microbiomeData::HMP_WGS, "WGS Species")
HMP_WGS_pathways <- getCollection(microbiomeData::HMP_WGS, "WGS Metagenome enzyme pathway abundance data" )

## get a correlation ComputeResult
## this is not necessarily to recommend pearson for functional data.
## it is simply exemplary. Always look at your data!
pathway_vs_species <- correlation(HMP_WGS_species, HMP_WGS_pathways, method = 'pearson')
```

## Interpreting Results

**Correlation Coefficients**: Assess the strength and direction of correlations. Positive coefficients 
indicate positive correlations, while negative coefficients indicate negative correlations.

**p-values and Adjusted p-values**: Identify biomarkers with statistically significant correlations, 
considering adjustments for multiple testing.

You can extract these metrics and sort and filter results by them:
```{r}
## you can extract network metrics 
pathway_vs_species.metrics <- as_tibble(
  getComputeResult(
    pathway_vs_species,
    correlationCoefThreshold = 0.5,
    pValueThreshold = 0.05
  )
)

## it's also easy to sort and filter these network metrics
## begin by renaming columns
colnames(pathway_vs_species.metrics) <- c('species', 'pathway', 'correlationCoef', 'pValue')
pathway_vs_species.metrics %>%
  filter(species == "Faecalibacterium prausnitzii") %>%
  filter(correlationCoef > 0.5) %>%
  arrange(desc(correlationCoef))
```

You can also visualize them with custom htmlwidgets:

```{r}
## now plot the network
## filters can be applied based on correlation coefficient and p-value
## renders an interactive htmlwidget, using a predetermined layout
correlationNetwork(
  pathway_vs_species,
  correlationCoefThreshold = 0.5,
  pValueThreshold = 0.05
)
```

Finally, for more advanced analysis of the network, you can extract it as an igraph object:

```{r}
## if you extract the network as an igraph object, you can get more detailed metrics
pathway_vs_species.igraph <- getComputeesult(pathway_vs_species, format = 'igraph')
degree <- degree(pathway_vs_species.igraph)
edgeBT <- edge_betweenness(pathway_vs_species.igraph)
pgRank <- page_rank(pathway_vs_species.igraph)
layout <- layout_with_kk(pathway_vs_species.igraph)

layout
```

